With rapid advances in quantum computing, many major technology companies have recently launched quantum computing cloud platforms, while smaller technology startups and research teams have released quantum simulation libraries for noisy intermediate-scale quantum (NISQ) devices. Notably, most quantum machines rely on gate-based architectures, in which quantum circuits form the backbone of these platforms. All quantum cryptography algorithms, ranging from cryptanalytic attack primitives to quantum key distribution protocols, must ultimately be executed as circuits. Because quantum computing suffers from decoherence, which degrades the integrity of information stored on quantum hardware, employing more quantum logic gates increases circuit depth and correspondingly reduces accuracy. Consequently, minimizing circuit depth remains a critical metric for improving cryptosystem performance on this class of architecture. In this work, we optimize the quantum circuits of symmetric-key block ciphers based on Shi and Feng's greedy algorithm. Our approach bidirectionally explores all possible shallow-depth CNOT circuits to synthesize the linear reversible components used in quantum cryptographic algorithms. We obtain a CNOT circuit for the AES MixColumns transformation with depth 9, outperforming previous greedy algorithms, and demonstrate that our method generalizes to other block ciphers. In addition, our algorithm is effective on a range of MDS matrices, synthesizing low-depth circuits for each. Furthermore, we refine several components of prior depth-oriented greedy algorithm research, achieving improved performance. Finally, we demonstrate the practical applicability of our approach to the KLEIN family, a lightweight block cipher that has recently attracted growing attention, by using our method to synthesize a resource-efficient, out-of-place MixNibbles component that reduces CNOT count by 64.4\% and circuit depth by 94.2\% relative to the original design, while still satisfying the required diffusion property.

\medskip
\textbf{Keywords:} Quantum circuit, Quantum depth, Greedy algorithm, Block cipher, AES MixColumns, Maximum Distance Separable Matrix, KLEIN